---
title: "nf-resolveome genotyping report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  donor_id: NULL
  ids: NULL
  wells: NULL
  plates: NULL
  cell_types: NULL
  min_alt_vaf: NULL
  min_alt_depth: NULL
---

```{r setup, include = F, message = F, warning = F, echo = F}
# libraries
library(magrittr)
library(tidyverse)
library(ggplot2)
library(pheatmap)

# chunk options
knitr::opts_chunk$set(warning = FALSE,
                      dpi = 300,
                      message = FALSE,
                      fig.align = "center")

# get n ids
n_ids <- length(params$ids)

# get cell annotations (missing columns are passed as "NA")
cells <-
  tibble(id = params$ids, well = params$wells, plate = params$plates,
         cell_type = params$cell_types) %>%
  mutate(across(c(well, plate, cell_type), ~ na_if(.x, "NA")))

n_plates <- n_distinct(cells$plate)

# label each cell by its cell type, plate (if more than one) and well
cells <-
  cells %>%
  mutate(cell_label = purrr::pmap_chr(
    list(cell_type, plate, well, id),
    function(cell_type, plate, well, id) {
      parts <- c(cell_type, if (n_plates > 1) plate, well)
      parts <- parts[!is.na(parts)]
      if (length(parts) == 0) id else paste(parts, collapse = " ")
    }))

# fall back to the id if the labels are not unique
if (anyDuplicated(cells$cell_label) > 0) {
  cells$cell_label <- cells$id
}

# can we annotate cells by their cell type?
annotate_cell_type <- !all(is.na(cells$cell_type))

# describe the filters
filters <- paste0("alt VAF >= ", params$min_alt_vaf,
                  ", alt depth >= ", params$min_alt_depth)
```

# Genotyped mutations

We load the annotated mutations.

```{r load_mutations}
muts <-
  read_tsv(paste0(params$donor_id, "_annotated_mutations.tsv"),
           col_types = cols(chr = col_character())) %>%
  alexr::type_variants() %>%
  mutate(mut_id = paste(chr, pos, ref, alt, sep = "_")) %>%
  left_join(cells, by = "id")
```

We filter to well-supported genotypes (`r filters`).

```{r filter_mutations}
geno <-
  muts %>%
  filter(alt_vaf >= params$min_alt_vaf, alt_depth >= params$min_alt_depth)
```

## Summary of genotyping

We summarise the genotyping.

```{r summarise_genotyping}
tibble(name = c("cells genotyped",
                "mutations genotyped",
                "mutations passing filters",
                "cells with >= 1 passing mutation"),
       value = c(n_ids,
                 n_distinct(muts$mut_id),
                 n_distinct(geno$mut_id),
                 n_distinct(geno$id))) %>%
  knitr::kable(caption = paste0("Summary of genotyping (", filters, ")."))
```

We plot the number of passing mutations per cell.

```{r plot_muts_per_cell, fig.width = 8, fig.height = max(5, 0.12 * n_ids)}
muts_per_cell <-
  cells %>%
  left_join(geno %>% dplyr::count(id, name = "n_muts"), by = "id") %>%
  mutate(n_muts = replace_na(n_muts, 0))

p_muts_per_cell <-
  muts_per_cell %>%
  ggplot(aes(y = reorder(cell_label, n_muts), x = n_muts))

if (annotate_cell_type) {
  p_muts_per_cell <- p_muts_per_cell + geom_col(aes(fill = cell_type), colour = "darkgrey")
} else {
  p_muts_per_cell <- p_muts_per_cell + geom_col(fill = "steelblue", colour = "darkgrey")
}

p_muts_per_cell +
  theme_classic() +
  theme(axis.text.y = element_text(size = 6), legend.position = "bottom",
        axis.title.y = element_blank()) +
  labs(x = "number of mutations", fill = "cell type") +
  scale_x_continuous(expand = c(0, 0))
```

## VAF heatmap

We plot the VAFs of the passing mutations across cells.

```{r prepare_heatmap}
p_dat <-
  geno %>%
  filter(!is.na(gene)) %>%
  mutate(mut_label = case_when(
    aachange == "." & impact == "Essential_Splice" ~ paste(chr, pos, "splice"),
    aachange == "." ~ paste(chr, pos, type),
    TRUE ~ aachange)) %>%
  mutate(mut_label = paste(gene, mut_label)) %>%
  # disambiguate any mutations that share a label
  group_by(mut_label) %>%
  mutate(mut_label = if (n_distinct(mut_id) > 1) {
    paste0(mut_label, " (", chr, ":", pos, ")")
  } else {
    mut_label
  }) %>%
  ungroup()

if (nrow(p_dat) > 0) {

  mat <-
    p_dat %>%
    pivot_wider(id_cols = mut_label, names_from = cell_label,
                values_from = alt_vaf, values_fn = max) %>%
    column_to_rownames("mut_label") %>%
    as.matrix()

  mat[is.na(mat)] <- 0

  # annotate each cell (column) with its cell type
  if (annotate_cell_type) {
    annot_col <-
      p_dat %>%
      distinct(cell_label, cell_type) %>%
      column_to_rownames("cell_label")
    annot_colors <- list(
      cell_type = setNames(
        scales::hue_pal()(n_distinct(annot_col$cell_type)),
        sort(unique(annot_col$cell_type))))
  } else {
    annot_col <- NA
    annot_colors <- NA
  }

} else {
  mat <- matrix(nrow = 0, ncol = 0)
}

# size the heatmap by its number of rows (mutations) and columns (cells),
# allowing space for the labels and dendrograms
hm_cell_size <- 0.14
hm_width <- min(40, max(8, hm_cell_size * ncol(mat) + 3))
hm_height <- min(40, max(6, hm_cell_size * nrow(mat) + 2))
```

```{r plot_heatmap, fig.width = hm_width, fig.height = hm_height}
if (nrow(mat) > 0) {

  pheatmap(
    mat,
    cluster_rows = nrow(mat) > 1,
    cluster_cols = ncol(mat) > 1,
    annotation_col = annot_col,
    annotation_colors = annot_colors,
    fontsize_row = 6, fontsize_col = 6,
    main = paste0(params$donor_id, " genotyping, ", filters))

} else {
  knitr::asis_output("No mutations passed the filters.")
}
```

## Mutations

We tabulate all passing mutations.

```{r display_mutations}
geno_display <-
  geno %>%
  select(id, chr, pos, ref, alt, gene, alt_vaf, alt_depth, total_depth,
         aachange, impact) %>%
  arrange(-alt_vaf, -alt_depth)

geno_display %>%
  knitr::kable(digits = 2,
               caption = paste0("All genotyped mutations (", filters, ")."))
```

We tabulate the cells carrying more than one mutation.

```{r display_multi_mutation_cells}
geno_display %>%
  add_count(id) %>%
  filter(n > 1) %>%
  arrange(-n, id) %>%
  knitr::kable(digits = 2,
               caption = paste0("Cells with multiple mutations (", filters, ")."))
```

We tabulate the mutations shared across multiple cells.

```{r display_shared_mutations}
geno_display %>%
  add_count(chr, pos, ref, alt) %>%
  filter(n > 1) %>%
  arrange(-n, chr, pos, ref, alt) %>%
  knitr::kable(digits = 2,
               caption = paste0("Mutations shared across multiple cells (",
                                filters, ")."))
```
