---
title: "nf-resolveome report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  donor_id: NULL
  ids: NULL
  seq_type: NULL
---

```{r setup, include = F, message = F, warning = F, echo = F}
# libraries
library(magrittr)
library(tidyverse)
library(ggplot2)

# chunk options
knitr::opts_chunk$set(warning = FALSE,
                      dpi = 300,
                      message = FALSE)

# get n ids
n_ids <- length(params$ids)
```

# Sequencing metrics

We load the coverage data.

```{r load_sequencing_metrics}
# initalise list of metrics
metrics <- list()

# get mosdepth file paths
mosdepth <-
  tibble(id = params$ids) %>%
  mutate(summary_txt = paste0(id, ".mosdepth.summary.txt"),
         dist = paste0(id, ".mosdepth.", ifelse(params$seq_type == "dna", "global", "region"), ".dist.txt"))

# load mean coverage
metrics$cov <-
  split(mosdepth$summary_txt, mosdepth$id) %>%
  purrr::map(read_tsv) %>%
  bind_rows(.id = "id") %>%
  dplyr::rename(chr = chrom) %>%
  filter(chr == "total" & params$seq_type == "dna" |
          chr == "total_region" & params$seq_type == "dnahyb") %>%
  transmute(id, name = "mean coverage", value = mean)

# load coverage distribution
dist <-
  split(mosdepth$dist, mosdepth$id) %>%
  purrr::map(read_tsv, col_names = c("chr", "cov", "prop"), col_types = "cdd") %>%
  bind_rows(.id = "id") %>%
  dplyr::filter(chr == "total")

# get proportion of genome covered per id
metrics$perc_cov <-
  dist %>%
  filter(cov == 1) %>%
  transmute(id,
            name = "% covered",
            value = 100 * prop)
```

We load the genotyped SNPs.

```{r heterozygosity}
# get genotyped snps
snps <-
  read_tsv(paste0(params$donor_id, "_genotyped_snps.tsv"),
           col_types = cols(chr = col_character()))
```

We calculate allelic dropout.

```{r allelic_balance}
# calculate allelic statuses
allelic_statuses <-
  snps %>%
  mutate(allelic_status = case_when(
    ref_depth > 0 & alt_depth > 0 ~ "both",
    ref_depth > 0 & alt_depth == 0 ~ "one",
    ref_depth == 0 & alt_depth > 0 ~ "one",
    ref_depth == 0 & alt_depth == 0 ~ "neither")) %>%
  dplyr::count(id, allelic_status) %>%
  group_by(id) %>%
  mutate(n_snps = sum(n), prop = n / n_snps,
         allelic_status = factor(allelic_status, levels = c("both", "one", "neither")))

# get prop allelic dropout
metrics$allelic_dropout <-
  allelic_statuses %>%
  filter(allelic_status %in% c("one", "neither")) %>%
  group_by(id) %>%
  summarise(name = "% allelic dropout", value = sum(prop) * 100)
```

We calculate the beta binomial overdispersion of the VAF distribution.

```{r beta_binomial}
fit_rho <- function(df) {
  tryCatch({
    # Log-likelihood for beta-binomial with fixed mu = 0.5
    ll <- function(rho) {
      rho <- plogis(rho)  # constrain to (0,1)
      -sum(VGAM::dbetabinom(df$alt_depth, df$total_depth, prob = 0.5, rho = rho, log = TRUE))
    }
    opt <- optim(0, ll, method = "Brent", lower = -10, upper = 10)
    tibble(rho = plogis(opt$par), n_snps = nrow(df))
  }, error = function(e) {
    tibble(rho = NA_real_, n_snps = nrow(df))
  })
}

metrics$overdispersion <-
  snps %>%
  group_by(id) %>%
  slice_sample(n = 50000, replace = FALSE) %>%  # subsample for speed
  group_modify(~ fit_rho(.x)) %>%
  ungroup() %>%
  transmute(id, name = "beta-binomial overdispersion", value = rho)
```

## Summary of metrics

We summarise all metrics.

```{r summarise_metrics}
# summarised
metrics %>%
  bind_rows() %>%
  group_by(name) %>%
  summarise(mean = mean(value), min = min(value), max = max(value)) %>%
  transmute(name,
            value = paste0(round(mean, 2), " (", round(min, 2), " - ", round(max, 2), ")")) %>%
  pivot_wider() %>%
  knitr::kable(caption = "Summary of metrics.")

# per cell
metrics %>%
  bind_rows() %>%
  tidyr::pivot_wider() %>%
  arrange(`mean coverage`) %>%
  knitr::kable(digits = 2, caption = "Metrics per cell.")
```

We plot all metrics.

```{r plot_metrics, fig.width = 12, fig.height = max(5, 0.15 * n_ids)}
blue_metrics <- c("mean coverage", "% covered")
metrics %>%
  bind_rows() %>%
  mutate(name = factor(name, levels = c("mean coverage", "% covered", "% allelic dropout", "beta-binomial overdispersion"))) %>%
  left_join(metrics$cov %>% transmute(id, cov = value)) %>%
  group_by(name) %>%
  mutate(
    fill_alpha = (value - min(value)) / (max(value) - min(value)),
    fill_col   = if_else(name %in% blue_metrics, "steelblue", "firebrick")
  ) %>%
  ungroup() %>%
  ggplot(aes(y = reorder(id, cov), x = value, fill = fill_col, alpha = fill_alpha)) +
  geom_col(position = "dodge", colour = "darkgrey") +
  ggh4x::facet_nested(~ name, scales = "free", space = "free_y") +
  theme_classic() +
  theme(strip.text.y.right = element_text(angle = 0), legend.position = "none",
        axis.title.x = element_blank(), axis.title.y = element_blank()) +
  scale_x_continuous(expand = c(0, 0), n.breaks = 3) +
  scale_fill_identity() +
  scale_alpha_continuous(range = c(0.15, 1)) +
  tidytext::scale_y_reordered() +
  guides(x = guide_axis(angle = -90, check.overlap = TRUE))
```

## Coverage distribution

We plot the coverage distribution.

```{r plot_coverage_dist}
dist %>%
  filter(prop >= 0.01) %>%
  ggplot(aes(x = cov, y = prop, colour = id)) +
  geom_line() +
  theme_classic() +
  labs(x = "coverage", y = "proportion of genome") +
  theme(legend.position = "none")
```

## Allelic balance

We plot the allelic balance of heterozygous SNPs.

```{r plot_allelic_vaf_dist, fig.height = 5, fig.width = 7}
alexr::plot_vaf_dist(snps,
                     paste(params$donor_id, "heterozygous SNPs"))
alexr::plot_vaf_dist(snps %>% filter(total_depth > 0),
                     paste(params$donor_id, "heterozygous SNPs (depth > 0)"))
```

## Allelic dropout

We plot the proportion of allelic dropout per cell.

```{r plot_allelic_dropout, fig.height = max(5, 0.12 * n_ids)}
allelic_statuses %>%
  left_join(
    allelic_statuses %>%
      filter(allelic_status == "both") %>%
      ungroup() %>%
      select(id, arr = prop)) %>%
  mutate(prop_lab = paste0(round(arr * 100) , "%")) %>%
  ggplot(aes(y = reorder(id, -arr), x = prop, fill = allelic_status)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("both" = "darkgreen", "one" = "orange", "neither" = "red")) +
  theme_classic() +
  theme(legend.position = "bottom", axis.text.y = element_text(size = 6)) +
  labs(y = "") +
  scale_x_continuous(expand = c(0, 0))
```